---
title: "Final Project Report"
author: "Ben Porter"
date: "12/15/2021"
output: github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ggplot2)
library(GGally)
library(mvShapiroTest)
library(DescTools)
library(reshape2)
library(biotools)
library(car)

sampledata <- read.csv("sample-data.csv")
sampledata1 <- sampledata

sampledata1$Date = as.factor(sampledata1$Date)
sampledata1$Continent = as.factor(sampledata1$Continent)
sampledata1$status = as.factor(sampledata1$status)



summary(sampledata1)

ggpairs(sampledata1[4:24])

```

a.

Research Question 1: 
What is the development status of a country dependent on?

Research Question 2: 
How many clusters can we separate our data into?

Research Question 3: 
Can we train a model to predict the development status of a country accurately based on our data?

c.
```{r}
# Box plots only for continuous variables
sampledata2 <- subset(sampledata1, select = c(5, 6, 7, 8, 12, 13, 14, 15, 16, 18, 20, 21, 22, 24))
summary(sampledata2)

col.continuous <- c(5, 6, 7, 8, 12, 13, 14, 15, 16, 18, 20, 21, 22)
qplot(status, value, data = melt(sampledata1, id.vars = 24, measure.vars = col.continuous),
      geom = "boxplot", na.rm = T) +
  facet_wrap("variable", nrow = 2, scales = "free_y") +
  labs(x = "result", y = NULL)
```

Variables of Interest Based on Research Questions and Significance:

Corruption.Index

Annual.GDP and GDP.per.capita

Debt and Debt.Per.Capita

Exports and Exports.per.GDP

Imports and Imports.per.GDP

Fertility.Rate

HDI

Life.expectancy

CO2.Tons.per.capita

status


f.
Data Visualization
```{r}
col.continuous <- c(5, 6, 7, 8, 12, 13, 14, 15, 16, 18, 20, 21, 22)
qplot(status, value, data = melt(sampledata1, id.vars = 24, measure.vars = col.continuous),
      geom = "boxplot", na.rm = T) +
  facet_wrap("variable", nrow = 2, scales = "free_y") +
  labs(x = "result", y = NULL)

ggpairs(sampledata2)
```



g.
Hierarchical Clustering
```{r}
library(fpc)
library(plyr)
library(reshape)
library(vegan)

sdata.euc.dist <- vegdist(sampledata2[1:13], method="euclidean")

sdata.euc.clust <- hclust(sdata.euc.dist, method="ward.D2")

# Plot within/between ratios against number of clusters
sdata.euc.ratio <- sapply(2:10, function(x) {
  cluster.stats(sdata.euc.dist, clustering = cutree(sdata.euc.clust, x))$wb.ratio
})
qplot(2:10, sdata.euc.ratio, geom = c("point", "line"),
      xlab = "Number of Clusters", ylab = "Within/Between Ratio", main = "Euclidean Distance (Ward's Method)")

# Plot Calinski-Harabasz index against number of clusters
sdata.euc.ch <- sapply(2:10, function(x) {
  cluster.stats(sdata.euc.dist, clustering = cutree(sdata.euc.clust, x))$ch
})
qplot(2:10, sdata.euc.ch, geom = c("point", "line"),
      xlab = "Number of Clusters", ylab = "ch Index", main = "Euclidean Distance (Ward's Method)")

sampledata2$cl.euc <- cutree(sdata.euc.clust, 5)

plotcluster(sampledata2[ ,1:13], sampledata2$cl.euc, method="dc")

```

A good number of clusters for this data is 5 because that is where the second elbow is located on the Within/Between Ratio graph and 5 is where the first peak is located on the ch Index.


h.
LDA
```{r}
sdata.lda <- lda(status ~ Annual.GDP + GDP.per.capita + Debt + Debt.Per.Capita + Corruption.Index + Exports + Exports...GDP + Imports + Imports...GDP + Fertility.Rate + HDI + Life.expectancy + CO2.Tons.per.capita,
                 data = sampledata2, prior = c(.15, .85), CV = F) 

predict(sdata.lda)$class

table(sampledata2$status, predict(sdata.lda)$class)

mean(sampledata2$status != predict(sdata.lda)$class)


```

Error rate is 7.24%. There are 31 miss-classified developed countries and 58 miss-classified developing countries.


QDA
```{r}
#qda
sdata.qda <- qda(status ~ Annual.GDP + GDP.per.capita + Debt + Debt.Per.Capita + Corruption.Index + Exports + Exports...GDP + Imports + Imports...GDP + Fertility.Rate + HDI + Life.expectancy + CO2.Tons.per.capita,
                 data = sampledata2, prior = c(.15, .85), CV = F)  

predict(sdata.qda)$class

table(sampledata2$status, predict(sdata.qda)$class)

mean(sampledata2$status != predict(sdata.qda)$class)  

```

Error rate is 6.26%. 18 miss-classified developed countries and 59 miss-classified developing countries.

QDA is better than LDA in terms of training error because it has a lower error rate.

i.

Test for Normality
```{r}
developing <- sampledata2[sampledata2[,14]=="Developing", -14]
developed <- sampledata2[sampledata2[,14]=="Developed", -14]

mvShapiro.Test(as.matrix(developing[1:13]))
mvShapiro.Test(as.matrix(developed[1:13]))

```

developing p-value: < 2.2e-16

developed p-value: < 2.2e-16

With all p-values < 0.05, we reject the null hypothesis, so each status is not from a multivariate normal distribution.

We should take caution when using MANOVA because of the lack of normality, but this does not make MANOVA invalid.



MANOVA
```{r}
fit.lm <- lm(cbind(Annual.GDP, GDP.per.capita, Debt, Debt.Per.Capita, Corruption.Index, Exports, Exports...GDP, Imports, Imports...GDP, Fertility.Rate, HDI, Life.expectancy, CO2.Tons.per.capita) ~ status, data = sampledata2)
summary(Manova(fit.lm))
```

Wilks test statistic: 0.3297165

F: 190.1551

p-value: 2.22e-16

p-value < 0.05, reject H0. At least two groups have different population means between the 13 variables.


PCA for developing countries
```{r}
developingpc <- prcomp(developing[1:13], scale = T, center = T)
summary(developingpc)

print(developingpc)

qplot(1:13, developingpc$sdev ^ 2, geom = c("point", "line"),
      xlab = "No. of PCs", ylab = "Component Variance (eigenvalue)", main = "Scree Plot")
```

3 PCs should be enough because that is where the first elbow is located on the scree plot and 3 PCs cover 74% of the variation.

PC1: The average of Annual.GDP, GDP.per.capita, Debt, Debt.Per.Capita, Corruption.Index, Exports, Exports...GDP, Imports, Imports...GDP, Fertility.Rate, HDI, Life.expectancy, CO2.Tons.per.capita minus Fertility.Rate for developed countries.

PC2: The average of Annual.GDP, Debt, Exports, and Imports minus the average GDP.per.capita, Debt.Per.Capita, Corruption.Index, Exports...GDP, and Imports...GDP for developing countries.

PC3: The average of Annual.GDP, HDI, Life.expectancy, and CO2.Tons.per.capita   minus the average Exports, Exports...GDP, Imports, Imports...GDP, and Fertility.Rate for developing countries.


PCA for developed countries
```{r}
developedpc <- prcomp(developed[1:13], scale = T, center = T)
summary(developedpc)

print(developedpc)

qplot(1:13, developedpc$sdev ^ 2, geom = c("point", "line"),
      xlab = "No. of PCs", ylab = "Component Variance (eigenvalue)", main = "Scree Plot")
```

3 PCs should be enough because that is where the first elbow is located on the scree plot and 3 PCs cover 75% of the variation.

PC1: The average of Annual.GDP, GDP.per.capita, Debt, Debt.Per.Capita, Corruption.Index, Exports, Imports, Fertility.Rate, HDI, Life.expectancy, CO2.Tons.per.capita, and Fertility.Rate minus Exports...GDP and Imports...GDP for developed countries.

PC2: The average of GDP.per.capita, Corruption.Index, Fertility.Rate, HDI, and Life.expectancy minus the average Annual.GDP, Debt, Debt.Per.Capita, Exports, Imports for developed countries.

PC3: The average of GDP.per.capita, Debt.Per.Capita, Exports, Exports...GDP, Imports, Imports...GDP, and HDI minus the average Life.expectancy for developed countries.

